\section{Author Contributions}

In this section we list the authors of this paper, grant support, and their contributions to this project, using the following \href{https://credit.niso.org/}{standard CRediT categories}:

\begin{enumerate}
    \item Conceptualization: Ideas; formulation or evolution of overarching research goals and aims.
    \item Data Curation: Management activities to annotate (produce metadata), scrub data and maintain research data (including software code, where it is necessary for interpreting the data itself) for initial use and later reuse.
    \item Formal Analysis: Application of statistical, mathematical, computational, or other formal techniques to analyze or synthesize study data.
    \item Funding Acquisition: Acquisition of the financial support for the project leading to this publication.
    \item Investigation: Conducting a research and investigation process, specifically performing the experiments, or data/evidence collection.
    \item Methodology: Development or design of methodology; creation of models.
    \item Project Administration: Management and coordination responsibility for the research activity planning and execution.
    \item Resources: Provision of study materials, reagents, materials, patients, laboratory samples, animals, instrumentation, computing resources, or other analysis tools.
    \item Software: Programming, software development; designing computer programs; implementation of the computer code and supporting algorithms; testing of existing code components.
    \item Supervision: Oversight and leadership responsibility for the research activity planning and execution, including mentorship external to the core team.
    \item Validation: Verification, whether as a part of the activity or separate, of the overall replication/reproducibility of results/experiments and other research outputs.
    \item Visualization: Preparation, creation and/or presentation of the published work, specifically visualization/data presentation.
    \item Writing - Original Draft Preparation: Creation and/or presentation of the published work, specifically writing the initial draft (including substantive translation).
    \item Writing – Review and Editing: Preparation, creation and/or presentation of the published work by those from the original research group, specifically critical review, commentary or revision – including pre- or post-publication stages.
\end{enumerate}

\begin{itemize}
    \item ...
    \item Pietro Monticone, Department of Mathematics, University of Trento, pietro.monticone@studenti.unitn.it: Data Curation, Formal Analysis, Project Administration, Resources, Software, Validation, Writing - Original Draft Preparation, Writing - Review and Editing.
    \item Terence Tao, Department of Mathematics, UCLA, tao@math.ucla.edu: Conceptualization, Data Annotation, Investigation, Methodology, Project Administration, Writing - Original Draft Preparation, Writing - Review and Editing. Supported by NSF grant DMS-2347850.
    \item Harald Husum, harald.husum@gmail.com: Investigation, Software, Visualization
    \item Jérémy Scanvic, Laboratoire de Physique, École Normale Supérieure de Lyon, jeremy.scanvic@ens-lyon.fr: Validation.
    \item ...
\end{itemize}
